Quantitative review of the effects of triazole and benzimidazole fungicides on Fusarium head blight and wheat yield in Brazil

### 2017-07-24 code.Rmd
Major update

---
title: "Meta-analysis fungicide efficacy FHB"
author: "Franklin Machado & Emerson Del Ponte"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: vignette
---

```{r, warning=FALSE, include=FALSE}

# Load packages 
library(tidyverse)
library(metafor)
library(ggthemes)
library(cowplot)
library(scales)
library(knitr)
library(nloptr)
library(minqa)
```

@@

@@

The data are stored in two csv files, one for FHB index and the other for yield. Each file contains only the selected treatments following a systematic review. See this table which contains the selected trials with summary information for the trial and the reference.

# import data

# FHB index data
fhb_sev <- read.csv("fhb_sev.csv", sep = ",", h = T)

# Yield data
fhb_yield <- read.csv("fhb_yield.csv", sep = ",", h = T)
```

## Prepare variables for meta-analysis

The most common form of effect size used in traditional meta-analysis are based on the contrasts of the treatment of interest with a common reference (e.g. control treatment), such as mean difference, response ratio, etc. This is known as the conditional modeling approach, also named contrast-based meta-analysis. An alternative and simpler approach, which is commonly used in plant pathology, is to fit a two-way linear mixed model directly to the treatment means. This is know as the uncondional modeling approach or arm-based meta-analysis.

@@

### Effect size calculations

# log of FHB index
fhb_sev$yi <- log(fhb_sev$sev)

# log of yield
fhb_yield$yi <- log(fhb_yield$yield)

# difference (D) in yield between treatment and non-treated check
fhb_yield$D <-fhb_yield$yield - fhb_yield$yield_check

### Sampling variance

# Sampling variance for the log of FHB index
fhb_sev$vi <- with(fhb_sev, V_sev / (n * sev^2))

@@

# Sampling variance for yield
fhb_yield$vi2 <- fhb_yield$V_yield/fhb_yield$n # multivariate approach

### Prepare the treatment variables

```{r}
# Check the number of entries by fungicide (AI: active ingredient) and number of sprays
table(fhb_sev$AI, fhb_sev$n_spray2)

# Create treatment variable based on fungicide name and number of sprays 
fhb_sev <- fhb_sev %>% 
  mutate(AI_nspray2 = paste(AI, n_spray2,sep='_')) %>% 
  filter(AI_nspray2 != "CARB_1") %>% 
  filter(trial, length(trial)>1)

@@

## ID variable for each entry within a trial

```{r}
fhb_sev$id <- 1:nrow(fhb_sev)
fhb_yield$id <- 1:nrow(fhb_yield)
```

## Descriptive analysis for yield

@@

# Number of trial by year
fhb_trial <- fhb_yield %>% 
  group_by(trial) %>% 
  filter(row_number() ==1)

@@

nrow(table(fhb_trial$location))#Ponta Grossa repete (??)

# Number of trial by publication
data.frame(table(fhb_trial$publication))

# Number of trial by state
data.frame(table(fhb_trial$state))

# entries with D value lower than zero
negat_D <- fhb_yield %>% 
  filter(D < 0)

@@

## Calculate coefficient of variation of FHB index and yield

```{r}
# CV for FHB index
## All entries
cv <- (sd(fhb_sev$sev)/mean(fhb_sev$sev))*100
cv

## Only non-treated check
cv <- (sd(fhb_sev$sev_check)/mean(fhb_sev$sev_check))*100
cv

## All entries
cv <- (sd(fhb_yield$yield)/mean(fhb_yield$yield))*100
cv

## For control
cv <- (sd(fhb_yield$yield_check)/mean(fhb_yield$yield_check))*100
cv
```

## Network meta-analysis

### Mean log of FHB index

@@

# Linear contrasts between pairs of treatments
anova(fhb_mv_AI, L=rbind(c(0,1,-1,0,0), 
                         c(0,0,-1,1,0),
                         c(0,0,-1,0,1),
                         c(0,0,0,-1,1)))

@@

## Control efficacy

Back-transform (exponential) the difference of the logs to obtain the response ratio and further calculate the percent reduction in FHB index relative to the check treatment (control efficacy) for each fungicide.

```{r}
results_AI <- data.frame(cbind((1-exp(fhb_mv_AI$b))*100, 
                              (1-exp(fhb_mv_AI$ci.lb))*100,
                              (1-exp(fhb_mv_AI$ci.ub))*100))
results_AI
```

## Inconsistency test

```{r, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

@@

### Model testing interaction treat x design
fhb_mv_AI_cons <- rma.mv(yi, vi, 
                         mods = ~ AI_nspray2 * cons_group, 
                         method="ML",
                         random = ~ AI_nspray2 | trial/cons_group, 
                         struct="UN", 
                         data=fhb_sev)

results_I
anova(fhb_mv_AI_cons, btt=11:20) #p-val = 0.9999

@@

## Moderator variables

### FHB index base

@@

### Yield base

```{r, message=FALSE, warning=FALSE}
# create binary variable (yield_class)
summary(fhb_sev$yield_check) # Median = 2993; Mean = 2915
fhb_sev$yield_class <- ifelse(fhb_sev$yield_check > 3000, 2, 1)

@@

### Year as continuous

@@

### Overall percent yield increase by treament

@@

### Overall absolute yield increase

@@

### Inconsistency test for yield data
```{r}

@@

## Visualize the data

###Boxplot of severity by treatment

```{r}
ggplot(fhb_sev, aes(AI_nspray2, sev))+
  geom_boxplot(size = 0.5, outlier.shape = NA)+ 
  geom_jitter(width=0.1, shape=1, size=2.5, color="gray50")+
  scale_x_discrete(labels=c("CHECK","CARB 2x", "PROP 2x" ,"TEBU 1x","TEBU 2x")) +
  theme_classic()+
  xlab("")+
  ylab("FHB index (%)")+ 
  scale_y_continuous(breaks=c(0, 10,  20,  30, 40, 50, 60, 70, 80, 90, 100), limits=c(0,100))

### Boxplot yield

ggplot(fhb_yield, aes(AI_nspray2, yield))+
  geom_boxplot(size = 0.5, outlier.shape = NA)+ geom_jitter(width=0.1, shape=1, size=2.5, color="gray50")+
  scale_x_discrete(labels=c("CHECK","CARB 2x", "PROP 2x" ,"TEBU 1x","TEBU 2x")) +
  theme_cowplot()+
  xlab("")+
  ylab("Yield (kg/ha)")+
  scale_y_continuous(breaks=c(0, 1000,  2000, 3000, 4000, 5000, 6000))

### Plot D

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(plyr)
fhb_yield$se <- sqrt(fhb_yield$vi2/sqrt(fhb_yield$n))

sum<-ddply(.data=fhb_yield, 
               .(AI_nspray2), 
               summarize, 
               n=paste("n =", length(AI_nspray2)))
sum1<-sum %>% filter(AI_nspray2 != "AACHECK_0", AI_nspray2!="PROP_2",AI_nspray2 !="TEBU_1",AI_nspray2 !="TEBU_2")

##### CARB ########

plot_D_carb <- fhb_yield %>% group_by(trial) %>% filter(AI_nspray2 != "AACHECK_0", AI_nspray2!="PROP_2",AI_nspray2 !="TEBU_1",AI_nspray2 !="TEBU_2") %>% 
  ggplot(., aes(reorder(treatment, D), D)) +
  geom_bar(stat="identity")+
  geom_errorbar(width=0.5, size= 0.3, aes(ymax = D + se, ymin=D - se))+
  geom_text(data=sum1, aes(x=5, y=1200,label=n), inherit.aes=FALSE, parse=FALSE)+
  geom_text(aes(x=5, y=1400,label="CARB 2x"), inherit.aes=FALSE, parse=FALSE)+
  geom_hline(aes(yintercept=0), size = 0.1)+
  scale_y_continuous(breaks=c(-250,0, 250,  500, 750, 1000, 1250, 1500))+
  ylab("D (kg/ha)")+ xlab("Study treatments (sorted) ") +
  theme(axis.text = element_text(size=11),
        #axis.title.x=element_blank(), 
        #axis.ticks.x=element_blank(),
        axis.text.x=element_blank(),
        legend.position="none")

######PROP######

sum2<-sum %>% filter(AI_nspray2 != "AACHECK_0", AI_nspray2!="CARB_2",AI_nspray2 !="TEBU_1",AI_nspray2 !="TEBU_2")

plot_D_prop <- fhb_yield %>% group_by(trial) %>% filter (AI_nspray2 != "AACHECK_0", AI_nspray2!="CARB_2",AI_nspray2 !="TEBU_1",AI_nspray2 !="TEBU_2") %>% 
  ggplot(., aes(reorder(treatment, D), D)) +
  geom_bar(stat="identity")+
  geom_errorbar(width=0.5, size= 0.3, aes(ymax = D + se, ymin=D - se))+
  geom_text(data=sum2, aes(x=4, y=1200,label=n), inherit.aes=FALSE, parse=FALSE)+
  geom_text(aes(x=4, y=1400,label="PROP 2x"), inherit.aes=FALSE, parse=FALSE)+
  geom_hline(aes(yintercept=0), size = 0.1)+
  scale_y_continuous(breaks=c(-250,0, 250,  500, 750, 1000, 1250, 1500))+
  ylab(" ")+ xlab("Study treatments (sorted) ") +
  theme(axis.text = element_text(size=11),
        axis.text.x=element_blank(),
        legend.position="none")

##### TEBU########

sum3<-sum %>% filter(AI_nspray2 != "AACHECK_0", AI_nspray2!="CARB_2",AI_nspray2 !="PROP_2",AI_nspray2 !="TEBU_2")

plot_D_tebu <- fhb_yield %>% group_by(trial) %>% filter (AI_nspray2 != "AACHECK_0", AI_nspray2!="CARB_2",AI_nspray2 !="PROP_2",AI_nspray2 !="TEBU_2") %>% 
  ggplot(., aes(reorder(treatment, D), D)) +
  geom_bar(stat="identity")+
 geom_errorbar(width=0.5, size= 0.3, aes(ymax = D + se, ymin=D - se))+
  geom_text(data=sum3, aes(x=5, y=1300,label=n), inherit.aes=FALSE, parse=FALSE)+
  geom_text(aes(x=5, y=1500,label="TEBU 1x"), inherit.aes=FALSE, parse=FALSE)+
  geom_hline(aes(yintercept=0), size = 0.1)+
  scale_y_continuous(breaks=c(-250,0, 250,  500, 750, 1000, 1250, 1500))+
  ylab("D (kg/ha)")+ xlab("Study treatments (sorted) ") +
  labs(fill="Number of sprays" )+
    theme(axis.text = element_text(size=11),
        axis.text.x=element_blank(),
        legend.position="top")

sum4<-sum %>% filter(AI_nspray2 != "AACHECK_0", AI_nspray2!="CARB_2",AI_nspray2 !="PROP_2",AI_nspray2 !="TEBU_1")

plot_D_tebu2 <- fhb_yield %>% group_by(trial) %>% filter (AI_nspray2 != "AACHECK_0", AI_nspray2!="CARB_2",AI_nspray2 !="PROP_2",AI_nspray2 !="TEBU_1") %>% 
  ggplot(., aes(reorder(treatment, D), D)) +
  geom_bar(stat="identity")+
  geom_errorbar(width=0.5, size= 0.3, aes(ymax = D + se, ymin=D - se))+
  geom_text(data=sum4, aes(x=5, y=1200,label=n), inherit.aes=FALSE, parse=FALSE)+
  geom_text(aes(x=5, y=1400,label="TEBU 2x"), inherit.aes=FALSE, parse=FALSE)+
  geom_hline(aes(yintercept=0), size = 0.1)+
  scale_y_continuous(breaks=c(-250,0, 250,  500, 750, 1000, 1250, 1500))+
  ylab(" ")+ xlab("Study treatments (sorted) ") +
  labs(fill="Number of sprays" )+
    theme(axis.text = element_text(size=11),
        axis.text.x=element_blank(),
        legend.position="top")

combo1 <- plot_grid(plot_D_carb, plot_D_prop,labels=c('', ''),  ncol = 2)

combo2 <- plot_grid(plot_D_tebu, plot_D_tebu2,labels=c('', ''),  ncol = 2)

plot_grid(combo1, combo2,labels=c('', ''),  ncol = 1)

>>>>

---
title: ''
output:
  html_document:
    css: my-style.css
    df_print: paged
    theme: flatly
    highlight: default
    toc: true
    toc_float: 
     collapsed: true
     smooth_scroll: false
---

@@

The most common form of effect size used in traditional meta-analysis are based on the contrasts of the treatment of interest with a common reference (e.g. control treatment), such as mean difference, response ratio, etc. This is known as the conditional modeling approach, also named contrast-based meta-analysis. An alternative and simpler approach, which is commonly used in plant pathology, is to fit a two-way linear mixed model directly to the treatment means. This is know as the unconditional modeling approach or arm-based meta-analysis.

@@

Let's first import the data and load the packages needed for the analyses.

```{r}
fhb_sev <- read.csv("fhb_sev.csv", sep = ",", h = T)
fhb_yield <- read.csv("fhb_yield.csv", sep = ",", h = T)
```

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Load packages 
library(tidyverse)
library(metafor)
library(ggthemes)
library(cowplot)
library(scales)
library(knitr)
library(nloptr)
library(minqa)
library(DT)
```

## Effect-sizes

@@

```{r}
# log of FHB index
fhb_sev$yi <- log(fhb_sev$sev)
hist(fhb_sev$yi)

# log of yield
fhb_yield$yi <- log(fhb_yield$yield)
hist(fhb_yield$yi)

# difference (D) in yield between treatment and non-treated check
fhb_yield$D <-fhb_yield$yield - fhb_yield$yield_check
hist(fhb_yield$D)
```

## Sampling variance

Let's create the sampling variance variables `vi` for the respective effect-sizes.

```{r}
# Sampling variance for the log of FHB index
fhb_sev$vi <- with(fhb_sev, V_sev / (n * sev^2))

@@

# Sampling variance for yield
fhb_yield$vi2 <- fhb_yield$V_yield/fhb_yield$n # multivariate approach
```

## Treatments

There are tree fungicides of interest in this study:

- Tebuconazole
- Propiconazole
- Carbendazim

Most of them were used as a single application but Tebuconazole was tested as single or sequential application. Let's get the number of entries per fungicide. The `AACHECK` refers to the plots where no fungicide was applied, or the control.

```{r}
table(fhb_sev$AI, fhb_sev$n_spray2)
```

Let's create another variable to identify each of these four treatments. We will omit CARB applied once because there are only three entries.

```{r}
library(tidyverse)
fhb_sev <- fhb_sev %>% 
  mutate(AI_nspray2 = paste(AI, n_spray2,sep='_')) %>% 
  filter(AI_nspray2 != "CARB_1") %>% 
  filter(trial, length(trial)>1)

@@

Finally, we create a unique identifier for the treatment within a trial. This can be used further as random effects in the model.

```{r}
# ID variable for each entry within a trial

fhb_sev$id <- 1:nrow(fhb_sev)
fhb_yield$id <- 1:nrow(fhb_yield)
```

## Descriptive analysis

We will get some summary information which may of use to describe the dataset when writing the material and methods and results section of the paper.

@@

# Number of trials by year
fhb_trial <- fhb_yield %>% 
  group_by(trial) %>% 
  filter(row_number() ==1)

@@

# Number of trials by publication type
data.frame(table(fhb_trial$publication))

# Number of trials by state
data.frame(table(fhb_trial$state))

# n. of entries with D value lower than zero
negat_D <- fhb_yield %>% 
  filter(D < 0)
```

We also want to check whether the difference in means (D) can be expressed in absolute rather than relative terms. The former is indicated for situations in which the coefficient of variation for the response variable in the check treatment is not too high (Paul et al. 2008)

```{r eval=FALSE, include=FALSE}
# CV for FHB index
## All entries
cv_index_all <- (sd(fhb_sev$sev)/mean(fhb_sev$sev))*100

## For control only
cv_index_check <- (sd(fhb_sev$sev_check)/mean(fhb_sev$sev_check))*100

## All entries
cv_yld_all <- (sd(fhb_yield$yield)/mean(fhb_yield$yield))*100

## For control only
cv_yld_check <- (sd(fhb_yield$yield_check)/mean(fhb_yield$yield_check))*100

@@

We can see that CV was high for FHB index, but much lower for yield. It is safer then to work with the absolute means difference for yield than FHB index, which will also require log-transformation.

## Meta-analysis

### Control efficacy

Let's fit an arm-based model using the `rma.mv` function of metafor to get the estimates for each treatment.

@@

@@

Let's see the summary ouput provided by this function. Note that the intercept is the control treatment and all treatments differed significantly from the control based on the very low P-values.

```{r}
fhb_mv_AI
```

We can set linear contrasts between treatments of interest and get the P-valued using the `anova` function.

```{r}
anova(fhb_mv_AI, L=rbind(c(0,1,-1,0,0), 
                         c(0,0,-1,1,0),
                         c(0,0,-1,0,1),
                         c(0,0,0,-1,1)))

Let's back-transform (exponential) the estimated mean (and respective 95%CIs) difference of the logs to obtain the response ratio, which is further expressed as percent reduction in FHB index relative to the check treatment (control efficacy) for each fungicide. We then prepare the data for plotting purposes.

```{r}
results_AI <- data.frame(cbind((1-exp(fhb_mv_AI$b))*100, 
                              (1-exp(fhb_mv_AI$ci.lb))*100,
                              (1-exp(fhb_mv_AI$ci.ub))*100))

treat <- c("Check", "CARB2x", "PROP2x", "TEBU1x", "TEBU2x")
efficacy <- tbl_df(results_AI)
efficacy$treat <- treat
efficacy$se <- fhb_mv_AI$se
colnames(efficacy) <- c("Mean", "CIsup", "CIinf", "Treatment", "SE")
efficacy
```

We can visualize this quickly, but we omit the check treatment.

```{r}
efficacy %>% 
  filter(Treatment != "Check") %>% 
  ggplot(aes(Treatment, Mean))+
  geom_point(aes(size=1/SE), shape=15)+
  geom_linerange(aes(ymin = CIinf, ymax = CIsup))+
  coord_flip()+
  theme_grey()+
  ylim(20,80)
```

#### Inconsistency: index

In a network meta-analysis, it is important to check whether the results are consistent across the different designs. We included this variable manually in the original dataset and tested the interaction effect of treatments and design group.

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

@@

### Model testing interaction treat x design
fhb_mv_AI_cons <- rma.mv(yi, vi, 
                         mods = ~ AI_nspray2 * cons_group, 
                         method="ML",
                         random = ~ AI_nspray2 | trial/cons_group, 
                         struct="UN", 
                         data=fhb_sev)
anova(fhb_mv_AI_cons, btt=7:10)

@@

#### Moderators

##### FHB index base

@@

##### Yield base

```{r, message=FALSE, warning=FALSE}

# create binary variable (yield_class)
summary(fhb_sev$yield_check) # Median = 2993; Mean = 2915
fhb_sev$yield_class <- ifelse(fhb_sev$yield_check > 3000, 2, 1)

@@

##### Year as continuous

@@

### Yield response: relative

@@

### Yield response: absolute

@@

#### Inconsistency

@@
